---
title: "Combining data for poverty mapping"
format: 
  html:
    fig-width: 9
    fig-height: 5
    fig-dpi: 300
editor: source
params:
  eval_all: FALSE
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE}
# Recollect data from the sources?
if ("try-error" %in% class(try({params$eval_all}, silent = TRUE)))
  params <- list(eval_all = FALSE, harvest = FALSE)
```

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
#| error: false

# Libraries
library(dplyr)
library(purrr)
library(tidyr)
library(readr)
library(pins)
library(readxl)
library(arrow)
library(glue)
library(stringr)
library(forcats)

library(ggplot2)
library(scales)
library(patchwork)

library(sf)
library(mapview)
library(flextable)
library(ftExtra)
library(knitr)
library(htmltools)
library(lubridate)

# Poland's database API
library(bdl)

# Chunk setup
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)

# Loading locally-developed
list.files("R", pattern="*.R$", full.names=TRUE, ignore.case=TRUE) |> 
  walk(source)

# Raw data root
root_raw <- "./data/raw"
root_temp <- "./data/temp"
root_clean <- "./data/clean"

# Data-storage boards
bd_raw <- root_raw |> file.path("api") |> board_folder(versioned = T)
bd_aux <- root_temp |> board_folder(versioned = T)
bd_clean <- root_clean |> board_folder(versioned = T)
```

# Loading the metadata

Metadata was prepared for each variable at the time of the raw data collection. It consist of metadata on variables groups (object `meta_groups`) and the variable-specific metadata (`meta_vars`). 

```{r eval=TRUE}
#| label: load-data-metadata
meta_groups <-
  root_raw |> file.path("pl-metadata.xlsx") |> readxl::read_xlsx(sheet = "pl") |> 
  fill(Category) |> 
  select(Category, id_indicator, Indicator, Source, Metadata, `Link to Source`)

meta_vars <- 
  root_raw |> file.path("pl-metadata.xlsx") |> readxl::read_xlsx(sheet = "pl-index") |> 
  fill(id_indicator, Indicator) |> 
  mutate(across(c(Indicator, var_name), ~ as_factor(.)))

geoms <- bd_aux |> pin_read("geometries")
```

## Preparing poverty ralated variables (TBD)

Here, poverty data should be loaded and visualized.

```{r}
# TBD
```

## Preparing area level variables

To prepare the area-level variables, one needs to have normalized data, where all variables are saved in a long-formatted table with four columns:

-   id: unique identifiers
-   year: year of the observation
-   var: variable code
-   val: value of the observation

Please note that IDs should be unique across all regions and levels. Therefore, filtering data by NUTS3 or LAU1 IDs will be possible later based on the corresponding mapping tables.

```{r echo=TRUE}
#| label: dta-example
all_dta <- bd_aux |> pin_read(name = "all_dta")
all_dta |> glimpse()
```

The, one needs to select a list of variables, which should selected for the analysis and provide it as a vector. Make sure that all variables are well documented in the metadata.

```{r}
var_focus <- c("gas_conn_resi", "foster_fam", "ntl_mean_mask",
               "elec_cons_percap", "tax_deduct_num", "fam_allow_num", 
               "ben_per10000", "empl_total", "pop_urbanize", "fam_allow_amt",
               "facil_res")
```

@tbl-nmissing reports the number of missing observations per administrative division level/year.

```{r}
#| label: tbl-nmissing
#| tbl-cap: N missing observations by year, variable and aggregation level
all_dta  |>
  filter(var %in% var_focus) |>
  get_missing(geoms = geoms, range = c(2011, 2018:2025)) |>
  mutate(across(where(is.character), ~ ifelse(is.na(.), "N/A", .))) |>
  add_varnames(meta = meta_vars) |>
  mutate(var_name = str_c("'", var, "': ", var_name)) |>
  select(var_name, level, matches("\\d{3,}")) |>
  arrange(var_name) |>
  group_by(var_name) |>
  as_flextable(hide_grouplabel = TRUE) |> 
  add_footer_lines(values = 
                     as_paragraph("Note: 'N/A' means that no data is available for the selected variable."))
```

Plots below plot the data on the map for all variables, key years, and key administrative levels.

```{r}
#| label: mass-plotting
var_focus |>
  walk(~ {
    try({
      out_plot <- all_dta |>
        plot_var_by_year(.x, geoms, meta_vars, years = c(2011, 2018, 2024))
      print(out_plot)
    }, silent = T)
  })
```

## Merging data

At the stage of data merging, some post-processing of the variables is also expected after the data is being converted to the wide format.

Example below shows how the merged data set could look like.

```{r echo=TRUE}
dta_rhs <-
  all_dta  |> filter(var %in% var_focus) |> 
  pivot_wider(names_from = var, values_from = val) |> 
  mutate(
    across(any_of(c("fam_allow_num", "foster_fam")),
           ~ log(.)),
    fam_allow_perfam =  fam_allow_amt / fam_allow_num
  )

dta_rhs |> glimpse()
```

To assure quality of newly created variables, they could be visually inspected.

```{r}
dta_rhs |>
  select(id, year, fam_allow_perfam) |>
  pivot_longer(fam_allow_perfam, names_to = "var", values_to = "val") |>
  plot_var_by_year("fam_allow_perfam", 
                   geoms = geoms, 
                   meta  = tibble(var_id = "fam_allow_perfam", 
                                  var_name = "Allowance per family", 
                                  var_units = "PLZ / Family")
                   , years = c(2011, 2018, 2024))
```

Resulting data set should be stored in the data-clean folder. It is advised using `pins` package as it can save the history of changes in the data. 

```{r saving data}
bd_clean |> pin_write(dta_rhs, name = "sae_data", type = "rds")
```
