---
title: "Preparing examplary data for poverty mapping"
format: 
  html:
    fig-width: 9
    fig-height: 5
    fig-dpi: 300
editor: source
params:
  eval_all: FALSE
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE}
# Recollect data from the sources?
if ("try-error" %in% class(try({params$eval_all}, silent = TRUE)))
  params <- list(eval_all = FALSE, harvest = FALSE)
```

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
#| error: false

# Libraries
pacman::p_load(
  pins, sf, data.table, tidyverse, car, msae, 
  sae, survey, spdep, knitr, MASS, caret)

# Chunk setup
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)

# Loading locally-developed
list.files("R", pattern="*.R$", full.names=TRUE, ignore.case=TRUE) |> 
  walk(source)

# Raw data root
root_raw <- "./data/raw"
root_temp <- "./data/temp"
root_clean <- "./data/clean-example"

# Data-storage boards
bd_raw <- root_raw |> file.path("api") |> board_folder(versioned = T)
bd_aux <- root_temp |> board_folder(versioned = T)
bd_clean <- root_clean |> board_folder(versioned = T)
```


## Geometries

Structure:

-   `id` integer or character id of the polygon. Unique across admin levels
-   `name` name of the admin level
-   `geometry` geometrical boundaries. Make sure to make them valid and relatively simple.

```{r}

# bd_clean |> pin_write(XXXX, name = "geometries", type = "rds")
```

# Example data

To prepare the area-level variables, (direct estimates of poverty, variances, all SE, and covariances, as well as the RHS variables, one needs to clean all this data to a single structure wide-format with following columns:

-   id: unique identifiers
-   year: year of the observation
-   var1, var2, var2, where these are variables with their unique, self-explaining names 

Please note that variable names should be documented somewhere. 

Please save data in the board `bd_clean` under two (or more) different pins, for example: 

-   `sea_data` for RHS variables
-   `pov_direct` for the direct estimates, variances, covariances, etc.

```{r eval=TRUE}
income_dt <- 
  root_raw |> 
  file.path("example-data", "incomedata_sample.RDS") |> 
  read_rds()
```

## Direct poverty estimates

```{r}

# bd_clean |> pin_write(df_out, name = "pov_direct", type = "rds")
```

## Right-hand side varaibles

```{r dt-lfs-load}

# bd_clean |> pin_write(XXXX, name = "sea_data", type = "rds")
```

